\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-04-26 Latex/main.tex
changes in part 2

For each narwhal, the entire track was split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time. We kept only GPS measures 24 
hours after tagging, to avoid any tagging effects on the narwhals's movement, as recommended in \cite{heide-jorgensen_behavioral_2021}.
Measures that resulted in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
Overall, $4192$ GPS positions were kept for the analysis and projected to UTM coordinates zone 26. The splitting between data before and after exposure resulted in $935$ measures before exposure and $3257$ measures after exposure. Table~\ref{table: data distribution} in Appendix shows how the data is distributed among the different narwhals. Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.

@@

For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure. We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing..

@@

		\node[above,orange] at (2.3,2.3) {$\hat{\Theta}_{ij}$};
	\end{tikzpicture}
	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain.}
	\label{fig: illustration nearest shore points}
\end{figure}
\begin{figure}[H]
    \centering
    \begin{subfigure}{0.48\textwidth}
    \centering
     \includegraphics[scale=0.45]{images/data_exploration/Dship_histo.png}
    \caption{Histograms of the distance to the ship \\ for each narwhal}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[scale=0.45]{images/data_exploration/Dshore_histo.png}
    \caption{Histograms of the distance to the shore \\ for each narwhal}
    \end{subfigure}
    \caption{Distance to ship and distance to shore covariates}
    \label{fig: distance to shore and distance to ship histograms}
\end{figure}

Exposure to ship covariate at time $t_{ij}$ for narwhal $i$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case. This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source. This provides a conservative estimate of the effect of the noise exposure. The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals. Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}. All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.

@@

	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity. This model is indeed a special case of hypoelliptic diffusion \cite{ditlevsen_hypoelliptic_2019}.
There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.

@@

By integrating the velocity equation over time one gets the following distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the covariance matrix $Q(\Delta)$ has an explicit form.

>>>>

For each narwhal, the entire track is split into a period before exposure defined as the period before the narwhal gets in line of sight with the ship for the first time, and a period "after" exposure, starting after the narwhal has been in line of sight with the ship for the first time. We keep only GPS measures 24 
hours after tagging, to avoid any tagging effects on the narwhals's movement, as recommended in \cite{heide-jorgensen_behavioral_2021}.
Measures that result in a velocity higher than $20$ km/h were also discarded (only 2 data points for the same narwhal track).
Overall, $4192$ GPS positions are kept for the analysis and projected to UTM coordinates zone 26. The splitting between data before and after exposure results in $935$ measures before exposure and $3257$ measures after exposure. Table~\ref{table: data distribution} in Appendix shows how the data is distributed among the different narwhals. Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure.

@@

For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure. We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.

@@

		\node[above,orange] at (2.3,2.3) {$\hat{\Theta}_{ij}$};
	\end{tikzpicture}
	\caption{Example of nearest shore point and angle $\Theta$. $\partial D$ represents the boundary of the domain, $y_{ij}$ is the observed GPS position of narwhal $i$ at time $t_{ij}$.}
	\label{fig: illustration nearest shore points}
\end{figure}
\begin{figure}[H]
    \centering
    \begin{subfigure}{0.48\textwidth}
    \centering
     \includegraphics[scale=0.45]{images/data_exploration/Dship_histo.png}
    \caption{Histograms of the distance to the ship \\ for each narwhal}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[scale=0.45]{images/data_exploration/Dshore_histo.png}
    \caption{Histograms of the distance to the shore \\ for each narwhal}
    \end{subfigure}
    \caption{Distance to ship and distance to shore covariates}
    \label{fig: distance to shore and distance to ship histograms}
\end{figure}

Exposure to ship covariate for narwhal $i$ at time $t_{ij}$ , denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case. This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source. This provides a conservative estimate of the effect of the noise exposure. The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals. Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}. All the relevant covariates that will be used for the analysis of narwhals movement are summarized in table \ref{tab: covariates}.

@@

	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.

@@

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the covariance matrix $Q(\Delta)$ has an explicit form.
